---
title: "Data Wrangling"
subtitle: "UW CSSCR"  
author: "Riddhi Mehta-Neugebauer <br> (based on slides by Charles Lanfear)"
date: "Jan 29, 2020 <br> Updated: `r format(Sys.Date(), format='%b %d, %Y')`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: tomorrow-night-bright
      highlightLines: true
      countIncrementalSlides: false
      titleSlideClass: ["center","top"]
---

```{r setup, include=FALSE, purl=FALSE}
options(htmltools.dir.version = FALSE)
```

# Overview

1. Subsetting Data

2. Creating Variables

3. Summarizing Data

4. Reshaping Data

---
# Packages Used Today

.smallish[
This workshop focuses on using packages from the [`tidyverse`](https://www.tidyverse.org/).
]

--

.smallish[
The `tidyverse` is a collection of R packages which share a design philosophy, syntax, and data structures.
]

--

.smallish[
The `tidyverse` includes the most used packages in the R world: [`dplyr`](https://dplyr.tidyverse.org/) and [`ggplot2`](https://ggplot2.tidyverse.org/)
]

--

.smallish[
You can install and load the *entire* `tidyverse` with the following:


```{r, eval = FALSE, warning=FALSE, message=FALSE}
install.packages("tidyverse")

```
```{r, warning=FALSE, message=FALSE}
library(tidyverse)
```
]

--

.smallish[
We will use two datasets. The first is the [`gapminder`](https://cran.r-project.org/web/packages/gapminder/README.html) dataset. And the second is the [`billboard`](https://cran.r-project.org/web/packages/billboard/index.html)


```{r, eval = FALSE, warning=FALSE, message=FALSE}
install.packages(c("gapminder", "billboard"))

```

```{r, warning=FALSE, message=FALSE}
library(gapminder); library(billboard)
```
]

---
# What's in gapminder?

Let's take a look at the variables in gapminder
```{r}

head(gapminder, 3)
```

---
class: inverse

# Modifying Data Frames with `dplyr`


---
# But First, Pipes (`%>%`)

`dplyr` uses the [`magrittr`](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html) forward pipe operator, usually called simply a **pipe**. We write pipes like `%>%` (`Ctrl+Shift+M`).

--

Pipes take the object on the *left* and applies it to the function on the *right*. 

--

```{r, message=FALSE, warning=FALSE}

gapminder %>% filter(country == "Canada") %>% head(2)
```

--

Why use pipes? They save us time, make code readable, and allow chaining like above, so we use them *all the time* when manipulating data frames.

---

# Using Pipes


Pipes are clearer to read when you have each function on a separate line (inconsistent in these slides because of space constraints).

--

```{r, eval=FALSE}
take_these_data %>%
    do_first_thing(with = this_value) %>%
    do_next_thing(using = that_value) %>% ...
```

--

Stuff to the left of the pipe is passed to the *first argument* of the function on the right. 

--

If you ever find yourself piping a function where data are not the first argument, use `.` in the data argument instead.
```{r, eval=FALSE}
gapminder %>% lm(pop ~ year, data = .)
```

---
# Pipe Assignment

When creating a new object from the output of piped functions, place the assignment operator at the beginning.

```{r, eval=FALSE}
lm_pop_year <- gapminder %>% 
  filter(continent == "Americas") %>%
  lm(pop ~ year, data = .)
```

No matter how long the chain of functions is, assignment is typically done at the top.

---

# `filter` Data Frames

I used **`filter()`** earlier. We subset *rows* of data using logical conditions with `filter()`!

```{r}
gapminder %>% filter(country == "Oman") %>% head(8)
```

What is this doing?

---

# Logical Operators

Logical operators test boolean conditions and return `TRUE`, `FALSE`, or `NA`.

--

`filter()` returns rows when the condition inside is `TRUE`.

We used `==` for testing "equals": `country == "Oman"`.

--

There are many other [logical operators](http://www.statmethods.net/management/operators.html):

--
* `!=`: not equal to

--
* `>`, `>=`, `<`, `<=`: greater than, greater than or equal to, etc.

--
* `%in%`: used with checking equal to one of several values

--

Or we can combine multiple logical conditions:

* `&`: both conditions need to hold (AND)

--
* `|`: at least one condition needs to hold (OR)

--
* `!`: inverts a logical condition (`TRUE` becomes `FALSE`, `FALSE` becomes `TRUE`)

---

# Multiple Conditions Example

Let's say we want observations from Oman after 1980 and through 2000.

--

```{r}
gapminder %>%
    filter(country == "Oman" &
           year > 1980 &
           year <= 2000 )
```

---
# `%in%` Operator

Common use case: Filter rows to things in some *set*.

We can use `%in%` like `==` but for matching *any element* in the vector on its right<sup>1</sup>. 

```{r}
former_yugoslavia <- c("Bosnia and Herzegovina", "Croatia", #<<
              "Macedonia", "Montenegro", "Serbia", "Slovenia") #<<

yugoslavia <- gapminder %>% filter(country %in% former_yugoslavia)
tail(yugoslavia, 2)
```

.footnote[[1] The `c()` function is how we make **vectors** in R, which are an important data type.]

--
Why don't you try an example?

---
## Sorting: `arrange()`

Along with filtering the data to see certain rows, we might want to sort it:

```{r}
yugoslavia %>% arrange(year, desc(pop))
```

The data are sorted by ascending `year` and descending `pop`.

---
## Selecting Columns: `select()`

Not only can we subset rows, but we can include specific columns (and put them in the order listed) using **`select()`**. 

```{r}
yugoslavia %>% select(country, year, pop) %>% head(4)
```

---
## Dropping Columns: `select()`


We can instead drop only specific columns with `select()` using `-` signs:

```{r}
yugoslavia %>% select(-continent, -pop, -gdpPercap) %>% head(4)
```

---
## Helper Functions for `select()`

.smallish[
`select()` has a variety of helper functions like `starts_with()`, `ends_with()`, and `contains()`, or can be given a range of continguous columns `startvar:endvar`. See `?select` for details. These are very useful if you have a "wide" data frame with column names following a pattern or ordering. 
]

.smallish[
```{r}
billboard %>% select(artist:wk3) %>% head(3)

```
]

--

.smallish[
```{r}
billboard %>% select(artist, track, num_range("wk", 10:13)) %>% head(3)
```
]

---
## Renaming Columns with `select()`


We can rename columns using `select()`, but that drops everything that isn't mentioned:

```{r}
billboard %>%
    select(artist, date = date.entered) %>%
    head(4)
```


---
### Safer: Rename Columns with `rename()`


**`rename()`** renames variables using the same syntax as `select()` without dropping unmentioned variables.

```{r}
yugoslavia %>%
    select(country, year, lifeExp) %>%
    rename(Life_Expectancy = lifeExp) %>%
    head(4)
```

---
class: inverse
# Creating Variables

---
## `mutate()`

In `dplyr`, you can add new columns to a data frame using **`mutate()`**.


```{r}
yugoslavia %>% filter(country == "Serbia") %>%
    select(year, pop, lifeExp) %>%
    mutate(pop_million = pop / 1000000, #<<
           life_exp_past_40 = lifeExp - 40) %>% #<<
    head(5)
```

.footnote[Note you can create multiple variables in a single `mutate()` call by separating the expressions with commas.]

---
# `ifelse()`


A common function used in `mutate()` (and in general in R programming) is **`ifelse()`**. It returns a vector of values depending on a logical test.

```{r, eval=FALSE}
ifelse(test = x==y, yes = this first_value , no = this second_value)
```

Output from `ifelse()` if `x==y` is...
* `TRUE`: `first_value` - the value for `yes =`

* `FALSE`: `second_value` - the value for `no = `

* `NA`: `NA` - because you can't test for NA with an equality!

--

For example:

```{r}
example <- c(1, 0, NA, -2)
ifelse(example > 0, "Positive", "Not Positive")
```

---
# `ifelse()` Example


```{r}
yugoslavia %>% mutate(short_country = 
                 ifelse(country == "Bosnia and Herzegovina", 
                        "B and H", as.character(country))) %>%
    select(short_country, year, pop) %>%
    arrange(year, short_country) %>%
    head(3)
```

Read the `ifelse()` part as "For each row in *yugoslavia* data, if country equals 'Bosnia and Herzegovina', make the new variable `short_country` equal to 'B and H', otherwise make it equal to that row's value of `country`."

This is a simple way to change some values but not others!

---
# `case_when()`

**`case_when()`** performs multiple `ifelse()` operations at the same time. `case_when()` allows you to create a new variable with values based on multiple logical statements. This is useful for making categorical variables or variables from combinations of other variables.

.smallish[
```{r}
gapminder %>% 
  mutate(gdpPercap_ordinal = 
    case_when(
      gdpPercap <  700 ~ "low",
      gdpPercap >= 700 & gdpPercap < 800 ~ "moderate",
      TRUE ~ "high" )) %>% # Value when all other statements are TRUE
  slice(6:9) # get rows 6 through 9
```
]

---
class: inverse

# Summarizing Data

---
## General Aggregation: `summarize()`

**`summarize()`** takes your column(s) of data and computes something using every row: 

* Count how many rows there are
* Calculate the mean
* Compute the sum
* Obtain a minimum or maximum value

You can use any function in `summarize()` that aggregates *multiple values* into a *single value* (like `sd()`, `mean()`, or `max()`).


---
# `summarize()` Example

For the year 1982, let's get the *number of observations*, *total population*, *mean life expectancy*, and *range of life expectancy* for former Yugoslavian countries.

```{r}
yugoslavia %>%
    filter(year == 1982) %>%
    summarize(n_obs = n(),
              total_pop = sum(pop),
              mean_life_exp = mean(lifeExp),
              range_life_exp = max(lifeExp) - min(lifeExp))
```

These new variables are calculated using *all of the rows* where `year` is equal to 1982 in `yugoslavia`

---
# Avoiding Repetition: 

### `summarize(across())`


Maybe you need to calculate the mean and standard deviation of a bunch of columns. With **`across()`**, put the variables to compute over first (using `c()` or `select()` syntax) and then put the functions to use in `list()` after.

```{r}
yugoslavia %>%
    filter(year == 1982) %>%
    summarize(across(c(lifeExp, pop), list(avg = ~mean(.), sd = ~sd(.))))
```

Note it automatically names the summarized variables based on names given in `list()`. 

---
# Avoiding Repetition

### Other functions:


There are additional `dplyr` functions similar to `summarize(across())`:

* `summarize_all()` and `mutate_all()` summarize / mutate *all* variables sent to them in the same way. For instance, getting the mean and standard deviation of an entire dataframe:

```{r, eval=FALSE}
yugoslavia %>% select(year, lifeExp, pop, gdpPercap) %>% summarize_all(funs(mean, sd))
```

* `summarize_if()` and `mutate_if()` summarize / mutate all variables that satisfy some logical condition. For instance, summarizing every numeric column in a dataframe at once:

```{r, eval=FALSE}
yugoslavia %>% summarize_if(is.numeric, funs(mean, sd))
```

You can use all of these to avoid typing out the same code repeatedly!

---
# `group_by()`


The special function **`group_by()`** changes how subsequent functions operate on the data, most importantly `summarize()`.

Functions after `group_by()` are computed *within each group* as defined by unique valus of the variables given, rather than over all rows at once.

Typically the variables you group by will be integers, factors, or characters, and *not continuous real values*.

---
# `group_by()` Example


```{r, warning=FALSE, message=FALSE}
yugoslavia %>%
  group_by(year) %>% #<<
    summarize(num_countries = n_distinct(country),
              total_pop = sum(pop),
              total_gdp_per_cap = sum(pop*gdpPercap)/total_pop) %>%
    head(4)
```

Because we did `group_by()` with `year` then used `summarize()`, we get *one row per value of `year`*!

Each value of year is its own **group**!

---
## Window Functions

Grouping can also be used with `mutate()` or `filter()` to give rank orders within a group, lagged values, and cumulative sums. You can read more about window functions in this [vignette](https://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html).

```{r}
yugoslavia %>% 
  select(country, year, pop) %>%
  filter(year >= 2002) %>% 
  group_by(country) %>%
  mutate(lag_pop = dplyr::lag(pop, order_by = year),
         pop_chg = pop - lag_pop) %>%
  head(4)
```

---
class: inverse
# Reshaping Data

---
# Slightly Messy Data

```{r, echo=FALSE}
enrollment <- 
  tibble(Program = c("Evans School", "Arts & Sciences",
                   "Public Health", "Other"),
       Female  = c(10, 5, 2, 5),
       Male    = c(6, 6, 3, 1))
pander::pander(enrollment, style="rmarkdown")
```


--

* What is an observation?
    + A group of students from a program of a given gender
* What are the variables?
    + Program, Gender, Count
* What are the values?
    + Program: Evans School, Arts & Sciences, Public Health, Other
    + Gender: Female, Male -- **one variable spread over two columns**
    + Count: **one variable spread over two columns**

---
# Tidy Version

```{r, message=FALSE, warning=FALSE}
(enrollment_long <- enrollment %>%
  pivot_longer(-Program, names_to="Gender", values_to="Count"))
```


* Each variable is a column.
* Each observation is a row.
* Ready to throw into `ggplot()` or a model!

---
# `tidyr` Functions

* **`pivot_longer()`**: takes a set of columns and pivots them down to make two new columns (which you can name yourself). 
    * Main arguments: 
        + `names_to=` is column that stores the original column names 
        + `value_to=` is column with the values in those original columns 
    * Both arguments must be in quotes--they don't exist yet!

--

* **`pivot_wider()`**: inverts `pivot_longer()` by taking two columns and pivoting them up into multiple columns. 
    * Main arguments:
        + `names_from=` column containing names of new columns 
        + `values_from=` column containing values for new columns
    * Arguments don't *need* to be in quotes--they already exist!

---
# Pivot Back

```{r}
(enrollment_wide <- enrollment_long %>%
  pivot_wider(names_from=Gender, values_from=Count))
```

This returns us to our original data format.

`pivot_wider()` is great for taking clean, tidy data and turning it into tables.

Wide format is also needed for certain estimation techniques such as structural equation models (e.g. `lavaan`).

---
# Resources

   * [UW CSSS508](https://clanfear.github.io/CSSS508/): Chuck Lanfear's University of Washington Introduction to R course which forms the basis for this workshop. All content including lecture videos is freely available.
   * [R for Data Science](http://r4ds.had.co.nz/) online textbook by Garrett Grolemund and Hadley Wickham. One of many good R texts available, but importantly it is free and focuses on the [`tidyverse`](http://tidyverse.org/) collection of R packages which are the modern standard for data manipulation and visualization in R.
   * [Advanced R](http://adv-r.had.co.nz/) online textbook by Hadley Wickham. A great source for more in-depth and advanced R programming.
   * [RStudio Primers](https://rstudio.cloud/learn/primers) for interactive, cloud-based tutorials.
   * [`swirl`](http://swirlstats.com/students.html): Interactive tutorials inside R.
   * [Useful RStudio cheatsheets](https://www.rstudio.com/resources/cheatsheets/) on R Markdown, RStudio shortcuts, etc.
